Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-08 draft.tex
added missing reference

Compared to the usual maximum-likelihood approach, we find for the method of moments:
\begin{enumerate}
  \item The angular observable ${S_i}$ can be determined
  independently of any other observable ${S_j}$. It is therefore
  much more robust to physics assumptions needed to define the full
  likelihood. In particular, this means one does not have to be
  specific regarding the form of new-physics contributions; in fact,
  one does not even need to be able to explicitly formulate the
  likelihood at all.
\item It is superior for a small number of samples $N$. Likelihood
  fits tend to be numerically unstable if lots of parameters need to
  be estimated from sparse data. This is more severe if the mode of
  the likelihood is near the boundary of the physically allowed
  region.\fred{Is there some LHCb note we could cite?}\danny{Ulrik suggests citing some
  statistical paper}
  For some of these decays of interest, there are only $\order{10}$--$\order{100}$ events
  per bin on record.\danny{What to cite? 1403.8004 only states total signal yield!}
\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
  smaller than those from the method of moments.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S5.png}
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S7.png}
        \caption{Uncertainty of the angular observables $S_5$ and $S_7$, extracted from $2\cdot 10^5$ studies of 
        simulated events for the decay $B\to K^*\ell^+\ell^-$. We show the uncertainty as a function of the number of simulated events $N$.
        The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
        \label{fig:errors}
\end{figure}

>>>>

Compared to the usual maximum-likelihood approach, we find for the method of moments:
\begin{enumerate}
  \item The angular observable ${S_i}$ can be determined
  independently of any other observable ${S_j}$. It is therefore
  much more robust to physics assumptions needed to define the full
  likelihood. In particular, this means one does not have to be
  specific regarding the form of new-physics contributions; in fact,
  one does not even need to be able to explicitly formulate the
  likelihood at all.
\item It is superior for a small number of samples $N$. Likelihood
  fits tend to be numerically unstable if lots of parameters need to
  be estimated from sparse data. This is more severe if the mode of
  the likelihood is near the boundary of the physically allowed
  region \cite{lehmann1998}. For some of these decays of interest, there are only $\order{100}$ events
  per bin on record.

 % \fred{Is there some LHCb note we could cite?}\danny{Ulrik suggests citing some
 % statistical paper}
 % For some of these decays of interest, there are only $\order{10}$--$\order{100}$ events
 % per bin on record.\danny{What to cite? 1403.8004 only states total signal yield!}
\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
  smaller than those from the method of moments.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S4.png}
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S8.png}
        \caption{Uncertainty of the angular observables $S_4$ (left) and $S_8$ (right), extracted from $2\cdot 10^5$ studies of 
        simulated events for the decay $B\to K^*\ell^+\ell^-$. We show the uncertainty as a function of the number of simulated events $N$.
        The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
        \label{fig:errors}
\end{figure}
